Statically Typed String Sanitation Inside a Python

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
updated

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
Web applications ultimately generate strings containing commands, which are   executed by systems like web browsers and database engines. Strings constructed from user input that has not been properly sanitized can thus cause command injection vulnerabilities. %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.

In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language and support operations like concatenation, substitution and coercion. Regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct such command strings. Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks. We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.

To be practical, adopting a type system like this should not require adopting a new programming language. Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.  We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, must control external systems that expose string-based command interfaces. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands contain data derived from user input, care must be taken to ensure that the user cannot provide an input that will subvert the intended command. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability if \verb|name| is controlled by a user: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, the program must \emph{sanitize} user input. 
For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that 
%escapes any double quotes, which are necessary in order to force uesr input to be interpreted as anything other than a string literal.
prepends double quotes and existing backslashes  with a backslash, which SQL treats safely. Note that this function is not idempotent, so it should only be called once. Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging.

We observe that most such sanitization techniques can be understood in terms of \emph{regular languages}. For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence consisting of characters other than quotation marks or backslashes, or escaped quotation marks or escaped backslashes. Concrete syntax like this can be understood to desugar, in a standard way, to the abstract syntax for regular expressions shown in Figure \ref{fig:regex}. We will work with this ``core'' for simplicity, and assume basic familiarity with regular expressions \cite{cinderella}.

In this paper, we present a static type system that tracks the regular language a string belongs to. We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}. These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been \emph{correctly} sanitized becomes manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). Run-time checks are only used when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated) and no additional space overhead is required. %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

Our type system for regular string types, $\lambdas$, is defined as an exntenstion to a simply-typed lambda calculus.
Opeartions on strings are given types according to corresponding, type-level
computations on regular expressions. A concatenation operation is defined in 
order to demosntrate this principle, but the premier operation of our system
is \emph{replacement}, which captures a rich set of input sanitation techniques
including escaped characters and encodings. In addition to defining this system,
we present a correctness-preserving compilation of $\lambdas$ into an underlying
language with ordinary strings as a regular expression library.

In addition to this theoretical contribution, we also explain how our extension
may be implemented in the Ace Programming Language, which is itself an extension
of Python -- a popular web programming language.

Summarily, we present a simple type system extension which ensures the absence
of input sanitation vulnerabilities by statically checking the correctness and 
correct use of input sanitation algorithms. This approach is \emph{composable}
in the sense that it is a conservative extension. This approach is \emph{complementary}
to exsiting input sanitation techniques that use regular expressions for input santiation.

@@

\subsection{Related Work and Alternative Approaches}

The input sanitation problem is well-understood. There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks. In this section, we explain how our approach to the input sanitation problem differs from each of these approaches. More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.

The Wyvern programming language provides a general framework for composing lanugage extensions \cite{tsl}\cite{wyvern}.
Our work identifies one particular extension, and is therefore complementary to Wyvern and related
work on extensible programming languages.
We are also unaware of any extensible programming languages which emphasize applications to security concerns.

In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.

@@

\section{A Type System for String Sanitation}

In this section we define a language for statically checked string sanitation ($\lambdas$).
The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
Expressions of this type evaluate to string literals in the language described by $r$.
Operations on expressions of type $\stringin{r}$ preserve this property.

The premier operation for manipulating strings in $\lambdas$ is string substitution, 
which is a familiar operation to any programmer who has used regular expressions.
The replacement operation replaces all instances of a pattern in one string with
another string; for instance, $\lsubst{a|b}{a}{c} = c$.
In order to compute the type resulting from aubstitution, we also need to compute
the result of replacing one language with another inside a given language.
Finally, just for convienance, we provide a coerce operation. The introduction of
coercion requires handling of runtime errors.

% Begin Grammar 
% Removes the < > from grammar productions.
\renewcommand{\grammarlabel}[2]{#1\hfill#2}

\begin{figure}[t]
\begin{grammar}
<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
$\stringin{r}$

<S> ::= 
      $\lambda x . e$ \hfill source terms \alt
      $e e$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{S}{S}$ \alt
      $\sisubst{r}{S}{S}$ \alt
      $\coerce{r}{S}$
\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
\end{grammar}
\end{figure}

\ 
\newpage
\ 
\newpage
\

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
\begin{mathpar}
%\inferrule[S-T-Abs]
%{\sctx, x : T_1 \vdash e : T_2}
%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[S-T-App]
%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
%{\sctx \vdash e_1 e_2 : T}
%  
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

@@

\inferrule[P-T-Check]
{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
{\tctx \vdash \tcheck{P_1}{P_2} : \str}
\end{mathpar}
\caption{Typing rules for our fragment of $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:etyping}
\end{figure}

@@

The main theorem of this paper is Theorem \ref{thm:main}. Establishing this result requires an additional theorem (Theorem \ref{tcorrect}), which establishes a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.

@@

\begin{thm}{Correctness of Input Sanitation for Translated Terms.}\label{thm:main}

@@

\section{Implementation in {Atlang}}
\begin{figure}
\begin{lstlisting}
#todo: why are these r's here?
#TODO-nrf this is maybe kind-of right bue I think there's a nicer way to do it.
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

In this section, we present an example atlang program using our extension together
with our implementation of the extension in Atlang modulo ${\sf lreplace}$.
Note that in addition to Figure \ref{fig:regex}, regular expressions in atlang
also support complementation using the carot.

TODO-nrf insert a pargraph or two which explains what's going on...

\subsection{An Example}

Figure \ref{fig:example} demosntrates the use of string types in atlang.
The sanitize function takes an arbitrary string and returns a string without
double quotes and left and right brackets. In this example, we use HTML
escape sequences; many variants of SQL provide their own syntax
for this as well.

TODO-nrf line by line of example.

\subsection{Extension Implementation}

TODO-nrf line by line of extension implementation as ewll.

An implementation of a similar extension as the one presented here is discussed in \cite{fulton12} and \cite{fulton13}.
Our prior work only supports a special case of replacement where unsafe substrings are completely removed.
In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
support for the replace operation as described in this paper.

@@

%$\psi$ & $\xlA{\psi}$ & \xA{\psi}\\

>>>>

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
Web applications must ultimately command systems like web browsers and database engines using strings. Strings constructed using user input that has not been properly sanitized can thus cause command injection vulnerabilities. %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.
In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language. These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct command strings. Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks. We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.

To be practical, adopting a specialized type system like this should not require the adoption of a new programming language. Instead, we favor extensible type systems: new type system fragments like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.  We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, control external systems by issuing  commands represented using  strings. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands must include data derived from user input, care must be taken to ensure that the user cannot construct input that subverts the intended command. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats as safely. Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging. Note that this function is not idempotent, so it should only be called once.

We observe that most such sanitization techniques can be understood in terms of \emph{regular languages} \cite{cinderella}. For example, \verb|name| should be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This concrete pattern syntax can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above. We write this as $\mathcal{L}\{r\}$. We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}. These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

We will begin in Sec. \ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambda_{RC}$. This allows us to specify the guarantees that the type system provides precisely. We also formally specify a translation from this calculus to a typed calculus with only standard strings and regular expressions, intending it as a guide to language implementors interested in building this feature into their own languages. This also demonstrates that no additional space overhead is required.

Waiting for a language designer to build this feature in is unsatisfying. Moreover, we would also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language requires empirically demonstrating that it is useful, but this is difficult to do if developers have no way to use it in practice. As such, we take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.

In Sec. \ref{atlang}, we show how to implement the type system fragment we have specified  using \texttt{atlang}, an extensible static type system implemented as a library inside Python. \texttt{atlang} leverages local type inference to control the semantics of literal forms, so reguar string types can be introduced using string literals without any run-time overhead. Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead. This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings.

We conclude after discussing related work in Sec. \ref{related}.

% TODO: need to update remainder of intro ... To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions. If $s : \stringin{r}$, then $s$ is a string matching the language $r$. Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$. The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$. Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted). We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.

% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library. This approach is \emph{composable} in the sense that it is a conservative extension. This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.

@@

@@

\section{Regular String Types, Minimally}\label{calculus}
\renewcommand{\grammarlabel}[2]{#1\hfill#2}

\begin{figure}[t]
\begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

\begin{figure}[t]
\begin{grammar}

<$\sigma$> ::=  $\sigma \rightarrow \sigma$     \hfill  source types          \alt
$\stringin{r}$

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x {:} \sigma . e$ \alt
      $e~(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\coerce{r}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
%\inferrule[S-T-Abs]
%{\sctx, x : T_1 \vdash e : T_2}
%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[S-T-App]
%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
%{\sctx \vdash e_1 e_2 : T}
%  
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Sec. \ref{translation}.

The syntax of $\lambda_{RS}$ is given in Figure \ref{fig:lamrs}, its static semantics are specified in Figure \ref{fig:etyping} and its evaluation semantics are specified in Figure \ref{fig:eeval}.

The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
Expressions of this type evaluate to string literals in the language described by $r$.
Operations on expressions of type $\stringin{r}$ preserve this property.

The premier operation for manipulating strings in $\lambdas$ is string substitution, 
which is a familiar operation to any programmer who has used regular expressions.
The replacement operation replaces all instances of a pattern in one string with
another string; for instance, $\lsubst{a|b}{a}{c} = c$.
In order to computer the type resulting from aubstitution, we also need to compute
the result of replacing on language with another inside a given language.
Finally, just for convienance, we provide a coerce operation. The introduction of
coercion requires handling of runtime errors.

% Begin Grammar 
% Removes the < > from grammar productions.

@@

\inferrule[P-T-Check]
{\tctx \vdash P_1 : \regex \\ \tctx \vdash P_2 : \str \\ }
{\tctx \vdash \tcheck{P_1}{P_2} : \str}
\end{mathpar}
\caption{Typing rules for our fragment of $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:ityping}
\end{figure}

@@

@@

\begin{thm}{Correctness of Input Sanitation for Translated Terms.}

@@

\section{Implementation as a Library in {Atlang}}\label{atlang}
\begin{figure}
\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') # TODO: is this right?
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

\section{Related Work}\label{related}
The input sanitation problem is well-understood. There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks. In this section, we explain how our approach to the input sanitation problem differs from each of these approaches. More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.

We are also unaware of any extensible programming languages which emphasize applications to security concerns.

In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
\section{Conclusion}

An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
support for the replace operation as described in this paper..

@@

%$\sigma$ & $\xlA{\sigma}$ & \xA{\sigma}\\
